---
title: "Final project"
author: Mandi, Marielena, Amy
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = FALSE, message = FALSE)
theme_set(theme_minimal())

library(tidyverse)
library(here)
library(rio)
library(janitor)
library(gtsummary)

```

### Description of the dataset and research questions

We will be using data from a study conducted in 2020 called ACT Suicide Prevention. This study collected dozens of variables of data, including responses to various assessments, such as the Beck Depression Inventory, from participants to determine if the ACT intervention for suicidality in college students was effective. Among the many variables gathered were participants’ age, living situation, race/ethnicity, gender, household income, and financial support. We are interested in looking at the relationship between college-aged female-identifying participants, household income, and levels of depression. Utilizing these four variables, gender, age, income, and severity of depression, we will be able to develop several visual graphs to help interpret the data and the relationships between variables, if any. Depending on our preliminary findings, we may decide to reduce or expand the variables we include in our project, as there are several available in the dataset.. Although there are no personal identifiers in the dataset, it is very personal information that has been gathered from participants. So we may need to “scramble” data for analysis and our final presentation. Our research question for this project will be “Is there a relationship between income rates in college-aged women and depression?” 
Variables of interest:  

- Age (dem 1.01)
- Gender (dem1.07a - demo1.07b)
- Income:  
  + Household income (dem1.13)
  + Income/job situation (dem1.14 - dem1.15b)
  + Financial support (dem1.16a - dem1.16g)
- Beck Depression Inventory (bdi1)

### Draft Data Script  

#### Load in data


```{r setup, include=FALSE}
act_t1 <- import(here("data", "ACT Suicide Prevention T1.sav"), setclass = "tbl_df") %>% 
  characterize() %>% 
  clean_names()

act_t1_12520 <- import(here("data", "ACT Suicide Prevention T1 1-25-20.sav"), setclass = "tbl_df") %>% 
  characterize() %>% 
  clean_names()

#I removed the duplicate chunk that was before the setup chunk
```

#### Conduct a full join to explore the differences in the two files shared with us. Both datasets had the same number of rows (one row per participant as confirmed using the `length()` and `unique()` functions), but one dataset had 137 columns and the other had 161 columns. We suspected the file with more columns contained additional variables. The joined file had no change in number of rows and 161 columns.

```{r}
dim(act_t1)
dim(act_t1_12520)
length(unique(act_t1$id))
length(unique(act_t1_12520$id))

act_all <- full_join(act_t1, act_t1_12520)

dim(act_all)

length(unique(act_all$id))
```

### Used select to pull out all columns with demographic infomation as well as BDI scores as these were the only variables we were intersted in looking at. From here we were able to pick which specific variables we wanted to look at relationships between.

```{r select}
act_all %>% 
  select(starts_with("dem1"), bdi1
    )
```

#### Calculate *M*, *SD*, *n*, min, and max of Beck Depression Inventory (BDI) scores by all reported genders.
```{r}
act_all %>% 
  group_by(dem1_07a) %>% 
  summarize(mean_bdi = mean(bdi1, na.rm = TRUE),
            sd_bdi = sd(bdi1, na.rm = TRUE),
            n_bdi = n(), 
            min_bdi = min(bdi1, na.rm = TRUE), 
            max_bdi = max(bdi1, na.rm = TRUE))

#It might be helpful to save these values using mutate() so that you have them appended to the dataset for future reference.
```

#### Calculate *M*, *SD*, and *n* of Beck Depression Inventory (BDI) scores by all reported genders and income level. This resulted in a table with columns for gender, income category, and descriptive stats for each group (rows corresponded to stats for a gender group and income group). We used `pivot_wider()` to widen the table by the gender variable. This resulted in  *M*, *SD*, and *n* columns for each gender. However, the resulting columns were not grouped by gender but by descriptive stat (e.g., mean_bdi_Man,mean_bdi_Transgender, mean_bdi_Woman). We wanted the descriptives (*M*, *SD*, and *n*) grouped by gender (e.g., *M*, *SD*, *n* for men, etc.). To accomplish this, we used `pivot_wider_spec()` to summarize BDI descriptive stats for each income category by reported gender. In other words, the resulting table has sets of columns of descriptives for each reported gender and one row per income category. We then used `pivot_longer()` to transform the table to a long form.

```{r}
act_all_descrip_gender_income <- act_all %>% 
  mutate(dem1_14 = factor(dem1_14, levels = c("Less than $5,000",
                                              "$5,000 - $9,999",
                                              "$10,000 - $14,999",
                                              "$15,000 - $19,999",
                                              "$20,000 - $24,999",
                                              "$25,000 - $29,999",
                                              "$50,000 or more",
                                              "I choose not to answer",
                                              "Unknown"
                                              )) )%>% 
  group_by(dem1_07a, dem1_14) %>% 
  summarize(mean_bdi = mean(bdi1, na.rm = TRUE),
            sd_bdi = sd(bdi1, na.rm = TRUE),
            n_bdi = n())

act_all_descrip_gender_income

act_all_descrip_gender_income_wide <- act_all_descrip_gender_income %>%
  pivot_wider(
    names_from = dem1_07a,
    values_from = c(mean_bdi, sd_bdi, n_bdi)
    ) %>% 
  arrange(dem1_14)

act_all_descrip_gender_income_wide

spec <- tibble(
  .name = c("mean_bdi_Man", 
            "sd_bdi_Man", 
            "n_bdi_Man", 
            "mean_bdi_Transgender", 
            "sd_bdi_Transgender", 
            "n_bdi_Transgender", 
            "mean_bdi_Woman", 
            "sd_bdi_Woman", 
            "n_bdi_Woman"),
  .value = c("mean_bdi", 
             "sd_bdi", 
             "n_bdi", 
             "mean_bdi", 
             "sd_bdi", 
             "n_bdi", 
             "mean_bdi", 
             "sd_bdi", 
             "n_bdi"),
  dem1_07a = c("Man",  
               "Man", 
               "Man",
               "Transgender", 
               "Transgender", 
               "Transgender", 
               "Woman",
               "Woman",
               "Woman")
)

spec

act_all_descrip_gender_income_wide_spec <- act_all_descrip_gender_income %>%
  pivot_wider_spec(spec) %>% 
  arrange(dem1_14)

act_all_descrip_gender_income_wide_spec

act_all_descrip_gender_income_long <- act_all_descrip_gender_income_wide_spec  %>%
  pivot_longer(
    cols = -dem1_14, 
    names_to = c("Descriptive Statistic", "Measure", "Reported Gender"),
    values_to = "Value",
    names_sep = "_",
  )

act_all_descrip_gender_income_long

#By creating all these different datasets, it will be hard to remember which dataset has the variables you are looking for. It may be helpful to just create one master dataset that you can work from.

```

#### We are interested in examining the relationship between income and depression (as measured with the BDI) for participants who identify as female. Filter dataset to only include female participants accordingly.

```{r}
act_all_female <- act_all %>% 
  filter(dem1_07a == "Woman")

#Are you interested in comparing gender groups or just looking at people who identify as female?
```

#### Explore the distribution of participants' ages. The majority of female participants are 18 years old. 

```{r}
act_all_female %>% 
  ggplot(aes(dem1_01)) +
  geom_bar(fill = "magenta") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Number of Female Participants by Age", y = "Number of Particpiants", x = "Age of Participants")

```

#### Distributions of BDI scores by age group. Distribution of BDI scores for 18-year-old female participants is positively skewed.

```{r}
act_all_female %>% 
  ggplot(aes(bdi1)) +
  geom_bar(fill = "magenta") + 
  facet_wrap(~dem1_01) +
  labs(title = "Rates of Depression among Female Participants by Age", subtitle = "Using results from the Beck Depression Inventory", y = "Number of Participants", x = "Rate of Depression")

```

#### Descriptive statistics of female participants' BDI scores by age

```{r}
act_all_female %>% 
  group_by(dem1_01) %>% 
  summarize(mean_bdi = mean(bdi1, na.rm = TRUE),
            sd_bdi = sd(bdi1, na.rm = TRUE),
            n_bdi = n(), 
            min_bdi = min(bdi1, na.rm = TRUE), 
            max_bdi = max(bdi1, na.rm = TRUE))

#Again, you can use mutate() to append these to the dataset
```

#### Side-by-side boxplots of BDI scores by age for all female participants 

```{r}
ggplot(act_all_female, aes(bdi1, dem1_01)) +
  geom_boxplot() + 
  coord_flip() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Rate of Depression among Female Participants by Age", subtitle = "Using responses to the Beck Depression Inventory", x = "Beck Depression Inventory Score", y = "Participant Age" )
```

#### Bar chart of personal gross annual income for all female participants 

```{r}
act_all_female %>% 
  mutate(dem1_14 = factor(dem1_14, levels = c("Less than $5,000",
                                              "$5,000 - $9,999",
                                              "$10,000 - $14,999",
                                              "$15,000 - $19,999",
                                              "$20,000 - $24,999",
                                              "$25,000 - $29,999",
                                              "$50,000 or more",
                                              "I choose not to answer",
                                              "Unknown"
                                              ))) %>% 
  ggplot(aes(dem1_14)) +
  geom_bar(fill = "darkgreen") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Female Participants by Annual Income", y = "Number of Participants", x = "Annual Income")
```

#### Side-by-side boxplots of BDI scores by personal gross annual income for female 18-year-old participants who were currently working

```{r}
act_all_female %>% 
  filter(dem1_01 == "18" | dem1_15a == "Yes") %>% 
  mutate(dem1_14 = factor(dem1_14, levels = c("Less than $5,000",
                                              "$5,000 - $9,999",
                                              "$10,000 - $14,999",
                                              "$15,000 - $19,999",
                                              "$20,000 - $24,999",
                                              "$25,000 - $29,999",
                                              "$50,000 or more",
                                              "I choose not to answer",
                                              "Unknown"
                                              ))) %>% 
  ggplot(aes(bdi1, dem1_14)) +
  geom_boxplot() + 
  coord_flip() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Rate of Depression in Female Participants by Annual Income", x = "Beck Depression Inventory Scores", y = "Annual Income")

```

#### Descriptive statistics (*M*, *SD*, *n*) of BDI scores for 18-year-old female participants who work by income bracket (group_by, summarize)

```{r}
act_all_female %>% 
  filter(dem1_01 == "18" | dem1_15a == "Yes") %>% 
  mutate(dem1_14 = factor(dem1_14, levels = c("Less than $5,000",
                                              "$5,000 - $9,999",
                                              "$10,000 - $14,999",
                                              "$15,000 - $19,999",
                                              "$20,000 - $24,999",
                                              "$25,000 - $29,999",
                                              "$50,000 or more",
                                              "I choose not to answer",
                                              "Unknown"
                                              ))) %>% 
  group_by(dem1_14) %>% 
  summarize(mean_bdi = mean(bdi1, na.rm = TRUE),
            sd_bdi = sd(bdi1, na.rm = TRUE),
            n_bdi = n(), 
            min_bdi = min(bdi1, na.rm = TRUE), 
            max_bdi = max(bdi1, na.rm = TRUE))
```

#### Trying out one of the table functions covered in Week 6 (gtsummary). The resulting table shows the counts and percentages of responses for each variable for female participants (e.g., number and percentage of female participants who reported their age as 17).

```{r}
act_all_female %>%
  gtsummary::tbl_summary()

#This table is very long and not very informative, perhaps you can collapse some of the rows so that there are not 1 datapoint per row.
```

#Great Job!

##Strengths

- Well done manupulating your data, I think you did a great job merging the data you had to create one cohesive dataset.
- The figures look very nice and well-labeled
-Good job using the pivot functions for your data to examine
 
##What I Learned

- There is a **strong** relationship between depression in female college students and income, in particular 18 year olds!
- 18 Year olds are more likely to complete a survey about income than other age groups.

##Areas to Improve

- Like I mentioned in my comments, it would be helpful to work from just one dataset rather than having multiple to work from
- The figures could be more informative. I think that they look quite nice but more information would drive home what your findings are.
- The table would benefit from collapsing the categories.